---
title: "My Notebook"
---
## our goal of collecting a dataset about liver disease is to facilitate research and analysis about it. since liver disease is more common and can be fatal we are aiming to classify the patients and analyze the data and correlations between the data.

link to the source of our dataset: https://www.kaggle.com/uciml/indian-liver-patient-records

general info about the dataset:
it has 11 columns which are:
 age (numeric)
 gender (symmetric binary)
 Total_Bilirubin (numeric)
 direct_Bilirubin (numeric)
 Alkaline_Phosphotase (numeric)
 Alamine_Aminotransferase (numeric)
 Aspartate_Aminotransferase (numeric)
 Total_Protiens (numeric)
 Albumin (numeric)
 Albumin_and_Globulin_Ratio (numeric)
 Dataset (asymmetric binary)

 and has 583 objects
 and class label (dataset) which indicates (1 disease or 2 no disease)


## The getwd() function tells you what the current working directory is
 setwd function to change the directory you work on 
```{r}
getwd()
setwd("C:/Users/arwa_/OneDrive/المستندات/GitHub/datamining-project")
```
 
 
## Importing the dataset
 read file named "indian_liver_patient" and save it in dataframe named "dataset"
```{r}
dataset = read.csv("C:/Users/arwa_/OneDrive/المستندات/GitHub/datamining-project/dataset/indian_liver_patient.csv")
```


## data analysis

 raw data and summary of the dataset
```{r}
View(dataset)
str(dataset)
summary(dataset)
```
we have 4 null values in Albumin_and_Globulin_Ratio column
we can also see there are some outliers

# Dealing with the missing values - Outliers 
# Replace missing Data with mean since there is only four tuples with a missing value
```{r}
dataset$Albumin_and_Globulin_Ratio = ifelse(is.na(dataset$Albumin_and_Globulin_Ratio), ave(dataset$Albumin_and_Globulin_Ratio, FUN =function(x) mean(x,na.rm=TRUE)), dataset$Albumin_and_Globulin_Ratio)
sum(is.na(dataset))
```

# stastical summaries
```{r}
summary(dataset$age)
summary(dataset$Total_Bilirubin)
summary(dataset$direct_Bilirubin)
summary(dataset$Alkaline_Phosphotase)
summary(dataset$Alamine_Aminotransferase)
summary(dataset$Aspartate_Aminotransferase)
summary(dataset$Total_Protiens)
summary(dataset$Albumin)
summary(dataset$Albumin_and_Globulin_Ratio)
```


## data visualization

# Create a scatterplot
```{r}
 plot(dataset$Alamine_Aminotransferase, 
     dataset$Aspartate_Aminotransferase, 
     xlab = "Alamine Aminotransferase", 
     ylab = "Aspartate Aminotransferase",
     main = "Scatterplot of Alamine Aminotransferase vs. Aspartate Aminotransferase",
     pch = 16, col = "blue")
```

 There is linear relationship (highly correlated) between Aspartate_Aminotransferase and Alamine_Aminotransferase. .  


# create histogram to see the number of male and femal patients
```{r}
hist(dataset$Alamine_Aminotransferase,
     col = "blue",           
     xlab = "ALT Levels",    
     ylab = "Frequency",      
     main = "Histogram of ALT Levels") 
```
 
# Create a bar plot
```{r}

```

## data preprocessing

# convert the gender value to numeric 
```{r}
dataset$Gender[dataset$Gender == "Male"] <- 0
dataset$Gender[dataset$Gender == "Female"] <- 1
```